---
title: "Density Dependent Expressions in popdemo and Compadre"
output: 
  html_document:
    theme: cosmo
    highlight: monochrome
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{popdemo: Demographic modelling using projection matrices}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
  
  
  ```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "##  ", collapse = TRUE, fig.align = "center")
library(popdemoDev)
options(digits = 4)
```

# Density dependent expressions
Density dependent expressions can be tricky to write, so we hope this will help
users figure out the correct specification for any density dependent model they 
may come across. It is possible to pass calls to `predict` into `makeMatExprs`, 
and this may save users considerable heartache. Examples of how to do that are below.


# Writing expressions for `makeMatExprs`
The ... argument of `makeMatExprs` takes named expressions. You do not need
to make a list of these before calling `makeMatExprs` (in fact, I'm not sure that
you could - R would try and fail to evaluate the symbols when generating the list!). 
Below are a few examples of typical models and how those would get re-written to
work in `makeMatExprs` and `project`.

### Density dependent survival

For this example, we'll consider a density dependent matrix model that has the following
vital rates:

$s = survival$

$f = fecundity$

$g = germination$

$V =  population\text{    } vector$



The matrix itself takes the form:

$$
\begin{bmatrix}
0 & 0 & f * g \\
s_1(V) & 0 & 0 \\
0 & s_2(V) & 0 
\end{bmatrix}
\begin{bmatrix}
V_1 \\
V_2 \\
V_3 \\
\end{bmatrix}

$$
The column vector to the right of the matrix is the population vector. Each entry in it represents the number of individuals in each age/stage category in the model.

Consider an example where $s_1$ depends on the sum of the population vector and $s_2$ depends on the product of the population vector. Survival probabilities are often modeled with logistic regressions. These take the form:

$s_i = \frac{1}{1 + e^{-(\beta_0 + \beta_1*x(V))}}$

Here, _s_ is the survival probability, $\beta_0$ is the intercept, $\beta_1$ is 
a slope, and _x(V)_ is some function the population vector _V_. In this case, we'll pretend 
_x(v)_ represents two different functions, so we'll need to write 4 different equations:
2 survival functions and 2 that calculate the quantities _x(V)_.

$s_1 = \frac{1}{1 + e^{-(\beta_0 + \beta_1*v_{sum})}}$

$s_2 = \frac{1}{1 + e^{-(\beta_0 + \beta_1*v_{prod})}}$

$v_{sum} = \Sigma_{i=1}^n V_i$

$v_{prod} = \Pi_{i=1}^n V_i$

where _i_ indexes all of the stages in the model and _n_ is the number of stages.

In this case, we need two expressions in a call to `makeMatExprs`. They should take
the following forms:

``` {r eval = FALSE}

mat_exprs <- makeMatExprs(
  s_1 = 1/(1 + exp(-(bs_0 + bs_1 * v_sum))),
  s_2 = 1/(1 + exp(-(bs_0 + bs_1 * v_prod))),
  
  v_sum = sum(v_1, v_2, v_3), 
  v_prod = v_1 * v_2 * v_3
)
```

Almost there! We also need to supply an expression that creates the matrix we want and an argument that specifies its dimension. Matrices are created by row internally.

``` {r eval = FALSE}
mat_exprs <- makeMatExprs(
  s_1 = 1/(1 + exp(-(bs_0 + bs_1 * v_sum))),
  s_2 = 1/(1 + exp(-(bs_0 + bs_1 * v_prod))),
  
  v_sum = sum(v_1, v_2, v_3), 
  v_prod = v_1 * v_2 * v_3,
  matrixExpression = c(
    0,   0,   f * g,
    s_1, 0,   0,
    0,   s_2, 0)
  ),
  matrixDimension = 3
)
```

Next, we generate a `dataList` object. In this case, we need to supply
initial values for $v_1$, $v_2$, $v_3$, $bs_0$, $bs_1$, $f$, and $g$. 
$v_1$, $v_2$, $v_3$ go into the final argument named `initPopVector`. Alternatively,
you can also just create those variables in your environment and `project()` will
find them there. 

``` {r eval = FALSE}

# With a call to makeDataList
data <- makeDataList(
  g = 0.05,
  f = 20,
  bs_0 = 0.5,
  bs_1 = 2.2,
  initPopVector = c(v_1 = 20, v_2 = 5, v_3 = 0)
)

# Or, just create the values in your environment. This is more likely useful when
# fitting your own matrix models, as parameters are typically created by manipulating
# other data first.

g <- 0.05
f <- 20
bs_0 <- 0.5
bs_1 <- 2.2

```

Now, we are ready to create our `CompadreDDM` object and iterate it!

```{r eval = FALSE}

# Using the dataList
myDDM <- CompadreDDM(dataList = data,
                     matExprs = mat_exprs)

myProjection <- project(myDDM)

# This will look a bit different as we don't supply an initial population vector.
# The default is then to use stage-biased vectors, so you should see 3 lines
# when you plot() the Projection

noDataDDM <- CompadreDDM(matExprs = mat_exprs)
noDataProjection <- project(noDataDDM) 


```

# Using `predict` in `makeMatExprs`
More complicated models can be difficult to write out (though we certainly do not want to discourage you from learning how to do so!). With that in mind, once your vital rate model is validated and looks good, it is possible to skip the above steps and substitute in a call to `predict` (provided that such a method exists for your model of choice. If not, you'll need to write the algebraic form). 

To start, fit your vital rate model. For this example, we'll do a fecundity model that depends on the combined density of the second and third stages. The matrix itself will take the following form:

$$
\begin{bmatrix}
0 & 0 & f(V) * g \\
s_1 & s_2 * (1-p) & 0 \\
0 & s_2*p & s_3 
\end{bmatrix}
\begin{bmatrix}
V_1 \\
V_2 \\
V_3 \\
\end{bmatrix}

$$

_s_'s represent survival probabilities, and for the sake of this example, are fixed. However, they don't
need to be - you can specify as many density dependent parameters as you like. _g_ is a germination probability, _p_ is a flowering probability. _f(V)_ is our density dependent fecundity.

```{r}
# simulate some data and then fit a model

fec <- rpois(100, 20:120)
densV2 <- round(runif(100, 20, 1000))
densV3 <- round(runif(100, 50, 700))

plotDens <- densV3 + densV2

# 
plotDens <- sort(plotDens, decreasing = TRUE)

fecModel <- glm(fec ~ plotDens, family = poisson())


# Now, we specify our constants. These aren't necessarily realistic values!
# We'll also use the stage-biased vector functionality of project() so there
# is no need to worry about the initPopVector argument

constants <- makeDataList(
  s_1 = 0.2,
  s_2 = 0.5,
  s_3 = 0.9,
  p = 0.6,
  g = 0.8
)

# Next are our matExprs. Here, we only need the density dependent terms,
# the matrix itself (matrixExpr), and matrix dimension (matrixDimension)

matExpressions <- makeMatExprs(
  f = predict(fecModel, 
              newdata = data.frame(plotDens = sum(V2, V3)),
              type = 'response'),
  matrixExpr = c(
    0, 0, f * g,
    s_1, s_2 * (1 - p), 0,
    0, s_2 * p, s_3
  ),
  matrixDimension = 3
)

fecDDM <- CompadreDDM(dataList = constants,
                      matExprs = matExpressions)

fecProjection <- project(fecDDM, vector = 'n', time = 50)

plot(fecProjection)

```

Before going any further, it is worth highlighting a couple common pitfalls. 

1.  Forgetting to specify the name of your predictor variable in the right hand side of `newdata = data.frame(**plotDens** = sum(V2,V3))`. Forgetting to specify this will result in errors reading like so: `Error in currentMat %*% evalEnv$popVec : non-conformable arguments` (the reason being that `predict` then returns a vector of predictions that's the same length as the original data!). 

2. Typos! These errors will have something like `Error in matrix(c(0, 0, rlang::eval_tidy(f) * g, s_1, s * (1 - p), 0,  : object 's' not found`. If these appear, carefully inspect both the `matExprs` and `dataList`.

Next, we'll add density dependent vital rate models for the _s_'s and _p_ and see how that plays out.


```{r}

# simulate some more data and model it
s_1 <- rbinom(100, 1, 0.3)
s_2 <- rbinom(100, 1, 0.5)
s_3 <- rbinom(100, 1, 0.9)
p <- rbinom(100, 1, 0.6)

# For realism(?)
plotDens <- sort(plotDens, decreasing = TRUE)

s1Mod <- glm(s_1 ~ plotDens, family = binomial())
s2Mod <- glm(s_2 ~ plotDens, family = binomial())
s3Mod <- glm(s_3 ~ plotDens, family = binomial())
pMod <- glm(p ~ plotDens, family = binomial())

# constants. Now, only one is needed (g)
constants <- makeDataList(g = 0.5)

# matrix expressions. This one will be a bit more verbose, but hopefully will
# make clear the power of this approach

matExpressions <- makeMatExprs(
  s_1 = predict(s1Mod, 
                newdata = data.frame(plotDens = sum(V2,V3)), 
                type = 'response'),
  s_2 = predict(s2Mod, 
                newdata = data.frame(plotDens = sum(V2,V3)), 
                type = 'response'),
  s_3 = predict(s3Mod, 
                newdata = data.frame(plotDens = sum(V2,V3)), 
                type = 'response'),
  p = predict(pMod, 
              newdata = data.frame(plotDens = sum(V2,V3)), 
              type = 'response'),
  f = predict(fecModel,
              newdata = data.frame(plotDens = sum(V2, V3)),
              type = 'response'),
  matrixExpr = c(0, 0, f * g,
                 s_1, s_2 * (1 - p), 0,
                 0, s_2 * p, s_3),
  matrixDimension = 3
)

fullDDM <- CompadreDDM(dataList = constants,
                       matExprs = matExpressions)

fullProjection <- project(fullDDM, vector = 'n', time = 75)

plot(fullProjection)

```

A bit chaotic, but hopefully this illustrates how to pass functions into vital rate expressions
with `CompadreDDM`s.

### Commonly used models and their translations to R code

For those that wish to specify their own models by hand rather than use `predict`,
here are a few examples that should help get started with the translations.

Poisson/Negative-binomial

$y = e^{\beta_0 + \beta_1*x +...}$

```{r eval = FALSE}
y = exp(b_0 + b_1 * x)

```


Linear model

$y = \beta_0 + \beta_1*x + ...$

```{r eval = FALSE}
y = b_0 + b_1 * x

```

Logistic regression

```{r eval = FALSE}

y = 1/(1 + exp(-(b_0 + b_1 * x)))
```


